---
title: "Exam 2 Review"
format: html
execute:
  eval: false
  warning: false
  message: false
format: 
  html:
    toc: true
    toc-depth: 3
---

## Additional helpful functions

-   `janitor::round_half_up(calculation, round)`

-   `MASS::fractions(calculation)`

### Using R

| Task                         | R code    | Example                       |
|------------------------------|-----------|-------------------------------|
| Save a value                 | `<-`      | `n <- 10`                     |
| Create a vector              | `c()`     | `x <- c(10, 20, 30)`          |
| Create consecutive intergers | `:`       | `0:5`                         |
| Sort values                  | `sort()`  | `sort(x)`                     |
| Addition / Subtraction       | `+` `-`   | `10 + 5` → `15``10 - 5` → `5` |
| Multiplication / Division    | `*` `/`   | `10 * 5` → `50``10 / 5` → `2` |
| Exponent / power             | `^`       | `4^2` → `16`                  |
| Square root                  | `sqrt()`  | `sqrt(25)` → `5`              |
| Absolute value               | `abs()`   | `abs(-4)` → `4`               |
| Grouping with Parentheses    | `()`      | `(10 + 5) / 3` → `5`          |

R requires `*` for multiplication, e.g., write `n * p` for $np$.

### Basic Functions in R

+--------------------+-------------------+------------------------------------------------+
| Calculation        | R code            | Example                                        |
+====================+===================+================================================+
| Sum                | `sum()`           | `sum(c(2, 4, 6))` → `12`                       |
+--------------------+-------------------+------------------------------------------------+
| Mean               | `mean()`          | `mean(c(2, 4, 6))` → `4`                       |
+--------------------+-------------------+------------------------------------------------+
| Median             | `median()`        | `median(c(2, 4, 9))` → `4`                     |
+--------------------+-------------------+------------------------------------------------+
| Min / Max          | `min()` `max()`   | `min(c(2, 4, 6))` → `2``max(c(2, 4, 6))` → `6` |
+--------------------+-------------------+------------------------------------------------+
| Variance           | `var()`           | `var(c(2, 4, 6))`                              |
+--------------------+-------------------+------------------------------------------------+
| Standard deviation | `sd()`            | `sd(c(2, 4, 6))`                               |
+--------------------+-------------------+------------------------------------------------+
| Number of values   | `length()`        | `length(c(2, 4, 6))` → `3`                     |
+--------------------+-------------------+------------------------------------------------+
| Round a result     | `round_half_up()` | `round_half_up(3.141, 2)` → `3.14`             |
+--------------------+-------------------+------------------------------------------------+
| Fraction           | `fractions()`     | `fractions(0.5)` → `1/2`                       |
+--------------------+-------------------+------------------------------------------------+
| Factorial          | `factorial()`     | `factorial(5)`                                 |
+--------------------+-------------------+------------------------------------------------+
| Combinations       | `choose()`        | `choose(5, 2)`                                 |
+--------------------+-------------------+------------------------------------------------+
| Summary statistics | `summary()`       | `summary(x)summary(data)`                      |
+--------------------+-------------------+------------------------------------------------+

## Recognizing the problem

+--------------------------------------------------------------------------------+----------------------------------------------------------+
| Clue in the problem                                                            | What to consider                                         |
+================================================================================+==========================================================+
| List all possible outcomes                                                     | Sample space                                             |
+--------------------------------------------------------------------------------+----------------------------------------------------------+
| Favorable outcomes out of equally likely outcomes                              | Basic probability                                        |
+--------------------------------------------------------------------------------+----------------------------------------------------------+
| A or B, or both                                                                | Union / addition rule                                    |
+--------------------------------------------------------------------------------+----------------------------------------------------------+
| A and B                                                                        | Intersection / multiplication rule                       |
+--------------------------------------------------------------------------------+----------------------------------------------------------+
| Given that B occurred                                                          | Conditional probability                                  |
+--------------------------------------------------------------------------------+----------------------------------------------------------+
| Events cannot occur together                                                   | Mutually exclusive events                                |
+--------------------------------------------------------------------------------+----------------------------------------------------------+
| One event does not change the other's probability                              | Independent events                                       |
+--------------------------------------------------------------------------------+----------------------------------------------------------+
| Selecting a group; order does not matter                                       | Combinations                                             |
+--------------------------------------------------------------------------------+----------------------------------------------------------+
| Without replacement from a finite population                                   | Dependent draws; hypergeometric for a count of successes |
+--------------------------------------------------------------------------------+----------------------------------------------------------+
| At least one                                                                   | Often easiest using the complement of zero               |
+--------------------------------------------------------------------------------+----------------------------------------------------------+
| Values with their probabilities                                                | Discrete random variable; weighted mean and variance     |
+--------------------------------------------------------------------------------+----------------------------------------------------------+
| All listed values are equally likely                                           | Discrete uniform                                         |
+--------------------------------------------------------------------------------+----------------------------------------------------------+
| Fixed number of independent trials, two outcomes, constant success probability | Binomial                                                 |
+--------------------------------------------------------------------------------+----------------------------------------------------------+
| Count of events in a time, space, or other interval at a constant average rate | Poisson                                                  |
+--------------------------------------------------------------------------------+----------------------------------------------------------+

## Probability rules

The **sample space** is the set of all possible outcomes. An **event** is a subset of those outcomes.

### Basic Probability

When individual outcomes are **equally likely**:

$$
P(A) = \frac{\text{number of favorable outcomes}}{\text{total number of outcomes}}
$$

```{r}
x / y
```

```{r}
favorable <- 3
total     <- 6

favorable / total
```

### Unions (A or B)

$$
P(A \cup B) = P(A) + P(B) - P(A \cap B)
$$

Subtract the overlap because it is counted in both $P(A)$ and $P(B)$. For mutually exclusive events, the overlap is zero.

```{r}
total  <- NA
n_A    <- NA
n_B    <- NA
n_both <- NA

(n_A + n_B - n_both) / total  
```

### Conditional Probability (A given B)

$$
P(A\mid B)=\frac{P(A\cap B)}{P(B)},\qquad P(B)>0.
$$

The condition changes the reference group to the outcomes in $B$.

The denominator is the count or probability for the event after the words **given that**. In general, $P(A\mid B)$ and $P(B\mid A)$ can differ.

Using counts from the same group, divide the number in both events by the number in the conditioning event.

```{r}
n_both <- 6
n_B    <- 14

n_both / n_B
```

### Intersections (A and B)

#### Independence

$$
P(A \cap B) = P (A\;and\;B) = P(A) \times P(B)
$$

```{r}
p_A <- 1 / 2
p_B <- 1 / 6

p_A * p_B
```

#### Dependent (General Multiplication Rule)

$$
P(A \cap B) = P (A\;and\;B)= P(A) \times P(B \vert A)
$$

```{r}
p_A  <- 5 / 8
p_BA <- 4 / 7

p_A * p_BA
```

### Complement

$$
P(A^c)= P (not\;A) = 1 - P(A)
$$

```{r}
p_A <- NA

1 - p_A
```

## Discrete Random Variable

List each possible numerical value with its probability. Probabilities must be between 0 and 1 and sum to 1.

#### Expected Value

The expected value is the long-run average over many repetitions. It does not have to be a possible value of $X$.

$$
E(X) = \sum_{i} x_i \, P(X = x_i)
$$

```{r}
# Possible values and their matching probabilities
x    <- c(   0,    1,    2)
prob <- c(0.20, 0.50, 0.30)

mu <- sum(x * prob)
mu
```

#### Variance and Standard Deviation

$$
\mathrm{Var}(X) = \sum_{i} (x_i - \mu)^2 \, P(X = x_i)\\\sigma = \sqrt{\mathrm{Var}(X)}
$$

```{r}
variance <- sum((x - mu)^2 * prob)
variance

standard_deviation <- sqrt(variance)
standard_deviation
```

## Uniform Distribution

Use a discrete uniform distribution when each listed outcome is equally likely.

$$
 P(X = x) = \frac{1}{n}
$$

```{r}
1 / n
```

#### Expected Value

$$
E(X) = \frac{\text{first value} + \text{last value}}{2}
$$

```{r}
(first +  last) / 2
```

### Variance

$$
 \operatorname{Var}(X)=\sigma^2=\frac{n^2-1}{12}
$$

```{r}
(n^2 - 1) / 12
```

## Factorials and Combinations

A factorial multiplies the positive integers up to $n$:

$$ n!=n(n-1)\cdots2(1),\qquad 0!=1. $$

When selecting $r$ objects from $n$ distinct objects **without regard to order**:

$$ \binom{n}{r}=\frac{n!}{r!(n-r)!}. $$

```{r}
factorial(n)
```

```{r}
choose(n, r)
```

## Binomial Distribution

Counts successes in $n$ trials: a fixed number of trials, two outcomes per trial, independent trials, and a constant success probability $p$. “Success” is the outcome being counted.

$$
 P(X = x) = \binom{n}{x} p^x (1-p)^{n-x}
$$

+----------------+-------------------+-------------------+
| Wording        | Mathematical Form | R Function        |
+================+===================+===================+
| exactly x      | $P(X=x)$          | `dbinom()`        |
+----------------+-------------------+-------------------+
| at most x\     | $P(X\le x)$       | `pbinom()`        |
| no more than x |                   |                   |
+----------------+-------------------+-------------------+
| less than x    | $P(X \lt x)$      | `pbinom(x-1)`     |
+----------------+-------------------+-------------------+
| more than x    | $P(X \gt x)$      | `1 - pbinom(x)`   |
+----------------+-------------------+-------------------+
| at least x     | $P(X \ge x)$      | `1 - pbinom(x-1)` |
+----------------+-------------------+-------------------+

| Argument | Equation | Meaning                              |
|----------|----------|--------------------------------------|
| `x`      | $x$      | number of successes                  |
| `size`   | $n$      | number of trials                     |
| `prob`   | $p$      | probability of success on each trial |

```{r}
dbinom(
  x    = NA, 
  size = NA,
  prob = NA
)
```

```{r}
pbinom(
  q    = NA, 
  size = NA,
  prob = NA
)
```

#### Expected Value

$$
E(X) = np
$$

```{r}
n * p
```

#### Variance and Standard Deviation

$$
Var(X) = np(1-p)\\SD(X) = \sqrt{np(1-p)}
$$

```{r}
n * p * (1 - p)
```

```{r}
sqrt(n * p * (1 - p))
```

## Poisson Distribution

Counts events in a fixed interval of time, length, area, or volume.

$$
P(X = x) = \frac{e^{-\lambda}\lambda^x}{x!}, \quad x = 0,1,2,\ldots
$$

where

-   $e \approx 2.71828$ \
    In R, we can use `exp(-lambda)`

-   $\lambda$ (lambda; Greek) = mean number of occurrences for the specified interval

-   $x$ = number of events

-   $x!$ (factorial) = means multiply all whole numbers from $x$ down to 1\
    $x! = x \times (x - 1) ... \times 1$\
    In R, we will use `factorial(a)`

For a Poisson distribution:

$$ \mu = \sigma^2 = \lambda $$

The mean equals the variance which equals $\lambda$. So, $\lambda$ controls **both** the **center** and **spread** of the distribution.

+----------------+-------------------+--------------------------+
| Wording        | Mathematical Form | R Function               |
+================+===================+==========================+
| exactly x      | $P(X=x)$          | `dpois(x, lambda)`       |
+----------------+-------------------+--------------------------+
| at most x\     | $P(X\le x)$       | `ppois(x, lambda)`       |
| no more than x |                   |                          |
+----------------+-------------------+--------------------------+
| less than x    | $P(X \lt x)$      | `ppois(x-1, lambda)`     |
+----------------+-------------------+--------------------------+
| more than x    | $P(X \gt x)$      | `1 - ppois(x, lambda)`   |
+----------------+-------------------+--------------------------+
| at least x     | $P(X \ge x)$      | `1 - ppois(x-1, lambda)` |
+----------------+-------------------+--------------------------+

| Argument | Equation  | Meaning |
|----------|-----------|---------|
| `x`      | $x$       | vector  |
| `lambda` | $\lambda$ | lambda  |

```{r}
(exp(-lambda) * lambda^x) / factorial(x)
```

```{r}
dpois(
  x      = NA,
  lambda = NA
)
```

```{r}
ppois(
  q      = NA,
  lambda = NA
)
```

## Hypergeometric Distribution

Counts successes in a random sample drawn **without replacement** from a finite population with known successes and failures. Every sample of the given size is equally likely.

$$
P(X = x) = \frac{\binom{k}{x}\binom{N-k}{\,n-x\,}}{\binom{N}{n}}
$$

where

-   $k$ = the total number of successes possible in the population

-   $N$ = the size of the total population

-   $n$ = the size of the sample drawn

-   $x$ = the number of successes in the sample of size $n$

+----------------+-------------------+-------------------+
| Wording        | Mathematical Form | R Function        |
+================+===================+===================+
| exactly x      | $P(X=x)$          | `dhyper()`        |
+----------------+-------------------+-------------------+
| at most x\     | $P(X\le x)$       | `phyper()`        |
| no more than x |                   |                   |
+----------------+-------------------+-------------------+
| less than x    | $P(X \lt x)$      | `phyper(x-1)`     |
+----------------+-------------------+-------------------+
| more than x    | $P(X \gt x)$      | `1 - phyper(x)`   |
+----------------+-------------------+-------------------+
| at least x     | $P(X \ge x)$      | `1 - phyper(x-1)` |
+----------------+-------------------+-------------------+

| Argument | Equation | Meaning                 |
|----------|----------|-------------------------|
| `x`      | $x$      | successes in sample     |
| `m`      | $k$      | successes in population |
| `n`      | $N-k$    | failures in population  |
| `k`      | $n$      | sample size             |

```{r}
choose(K, x) * choose(N - K, sample_size - x) /
  choose(N, sample_size)
```

```{r}
dhyper(
  x = NA, # number of successes in sample
  m = NA, # number of successes in population
  n = NA, # number of failures in population
  k = NA  # sample size
)
```

```{r}
phyper(
  q = NA, # number of successes in sample
  m = NA, # number of successes in population
  n = NA, # number of failures in population
  k = NA  # sample size
)
```

#### Expected Value

The expected value tells us the average number of successes we would expect if we repeated this sampling process many times.

$$
E(X) = n \cdot \frac{k}{N}
$$

```{r}
n * (k / N)
```

```{r}
population_size <- 15
successes_pop   <- 4
sample_size     <- 2

expected_value <- sample_size *
  (successes_pop / population_size)

expected_value
```

#### Variance

$$
\operatorname{Var}(X)=n \cdot \frac{k}{N}\left(1 - \frac{k}{N}\right)\frac{N - n}{N - 1}
$$

```{r}
sample_size *
  (successes_pop / population_size) *
  (1 - successes_pop / population_size) *
  ((population_size - sample_size) /
     (population_size - 1))
```

```{r}
population_size <- NA
successes_pop   <- NA
sample_size     <- NA

variance <- sample_size *
  (successes_pop / population_size) *
  (1 - successes_pop / population_size) *
  ((population_size - sample_size) /
     (population_size - 1))

variance
```
