---
title: "Exam 1 Formula Guide"
format: 
  html: 
    toc: true
    toc-location: left
    toc-title: "Questions"
    toc-depth: 2
    css: styles.css
engine: knitr
execute:
  eval: false
  warning: false
  message: false
---

## Suggestions

-   Classify the problem to determine which calculation or formula you need.

-   Careful writing down the data! Watch out for [typos]{.underline}.\
    Pasting tables into Excel might paste pictures (whomp whomp).

-   Round to the specified number of digits [at the last calculation]{.underline}.

-   Be [strategic]{.underline} with your available time.

-   **YOU GOT THIS!**

### Using R

| Task            | R code   | Example              |
|-----------------|----------|----------------------|
| Create a vector | `c()`    | `x <- c(10, 20, 30)` |
| Save a value    | `<-`     | `n <- 10`            |
| Sort values     | `sort()` | `sort(x)`            |

### Basic Math in R

+----------------------+----------------------+----------------------+
| Operation            | R code               | Example              |
+======================+======================+======================+
| Addition             | `+`                  | `10 + 5` → `15`      |
+----------------------+----------------------+----------------------+
| Subtraction          | `-`                  | `10 - 5` → `5`       |
+----------------------+----------------------+----------------------+
| Multiplication       | `*`                  | `10 * 5` → `50`      |
+----------------------+----------------------+----------------------+
| Division             | `/`                  | `10 / 5` → `2`       |
+----------------------+----------------------+----------------------+
| Exponent / power     | `^`                  | `4^2` → `16`         |
+----------------------+----------------------+----------------------+
| Square root          | `sqrt()`             | `sqrt(25)` → `5`     |
+----------------------+----------------------+----------------------+
| Absolute value       | `abs()`              | `abs(-4)` → `4`      |
+----------------------+----------------------+----------------------+
| Parentheses          | `()`                 | `(10 + 5) / 3` → `5` |
+----------------------+----------------------+----------------------+

### Basic Functions in R

+---------------------+----------------------------+-----------------------------------------------+
| Calculation         | R code                     | Example                                       |
+=====================+============================+===============================================+
| Sum                 | `sum()`                    | `sum(c(2, 4, 6))` → `12`                      |
+---------------------+----------------------------+-----------------------------------------------+
| Mean                | `mean()`                   | `mean(c(2, 4, 6))` → `4`                      |
+---------------------+----------------------------+-----------------------------------------------+
| Median              | `median()`                 | `median(c(2, 4, 9))` → `4`                    |
+---------------------+----------------------------+-----------------------------------------------+
| Minimum             | `min()`                    | `min(c(2, 4, 6))` → `2`                       |
+---------------------+----------------------------+-----------------------------------------------+
| Maximum             | `max()`                    | `max(c(2, 4, 6))` → `6`                       |
+---------------------+----------------------------+-----------------------------------------------+
| Range               | `range()`                  | `range(c(2, 4, 6))` → `2 6`                   |
+---------------------+----------------------------+-----------------------------------------------+
| Variance            | `var()`                    | `var(c(2, 4, 6))`                             |
+---------------------+----------------------------+-----------------------------------------------+
| Standard deviation  | `sd()`                     | `sd(c(2, 4, 6))`                              |
+---------------------+----------------------------+-----------------------------------------------+
| Number of values    | `length()`                 | `length(c(2, 4, 6))` → `3`                    |
+---------------------+----------------------------+-----------------------------------------------+
| Round a result      | `janitor::round_half_up()` | `janitor::round_half_up(3.14159, 2)` → `3.14` |
+---------------------+----------------------------+-----------------------------------------------+
| Summary statistics  | `summary()`                | `summary(x)`                                  |
|                     |                            |                                               |
|                     |                            | `summary(data)`                               |
+---------------------+----------------------------+-----------------------------------------------+

### Excel Formulas

-   =TEXTJOIN(", ", TRUE, A1:A14)

-   =SUM()

-   Simple math ( + - / etc.)

-   Lock a cell reference with \$ (e.g., \$A\$1)

## Frequency Tables

Relative frequency is:

$$
\text{Relative Frequency} = \frac{f}{n}
$$

Percentage is:

$$
\text{Percentage} = \frac{f}{n}\times 100
$$

where $f$ is the frequency and nn is the total number of observations.

```{r}
install.packages("descr") #install the package *once* if you chose to use this option
library(descr) # load the package from your library
descr::freq(as.ordered(data$variable), plot = FALSE) # function freq()
```

```{r}
x <- c("A", "A", "B", "B", "B", "C")

freq <- table(x)
n <- length(x)

freq                    # Frequency
freq / n                # Relative frequency
(freq / n) * 100        # Percentage
cumsum(freq)            # Cumulative frequency
```

## Central Tendency

#### Mode

```{r}
library(lsr)
```

```{r}
lsr::modeOf(data$variable, na.rm = TRUE)
```

```{r}
# Simple histogram
x <- c(13,13,-5,-5,13,11,-5)
hist(x, breaks = 10)           # breaks = bins = classes
```

#### Median

```{r}
x <- c(13,13,-5,-5,13,11,-5)
median(x)
```

```{r}
median(data$variable, na.rm = TRUE)
```

#### Mean

The arithmetic mean is:

$$
\bar{x} = \frac{\sum x_i}{n}
$$

where $\bar{x}$ is the sample mean, $x_i$ represents each observation, and $n$ is the number of observations.

```{r}
(56 + 31 + 56 + 8 + 32) / 5
```

```{r}
x <- c(13,13,-5,-5,13,11,-5)
sum(x) / length(x)
```

```{r}
x <- c(56, 31, 56, 8, 32)
mean(x, na.rm = TRUE)
```

#### Weighted Mean

The weighted mean is:

$$
\bar{x}_w = \frac{\sum w_i x_i}{\sum w_i}
$$

where $x_i$ is each observation and $w_i$ is its corresponding weight.

```{r}
# Weighted Mean
x <- c(24, 41, 58, 75, 92) # observations
w <- c(2, 8, 15, 8, 11)    # weights

sum(x * w) / sum(w)
```

```{r}
# Weighted Mean Function
x <- c(24, 41, 58, 75, 92) # observations
w <- c(2, 8, 15, 8, 11)    # weights

weighted.mean(x, w, na.rm = TRUE)
```

The weighted sample variance is:

$$
s^2 = \frac{\sum w_i(x_i-\bar{x}_w)^2}{n-1}
$$

where $n = \sum w_i$.

```{r}
# variance, calculate after calculating the mean
x <- c(24, 41, 58, 75, 92) # observations
w <- c(2, 8, 15, 8, 11)    # weights

n <- sum(w)

xbar <- sum(x * w) / n                   # weighted mean
s2   <- sum(w * (x - xbar)^2) / (n - 1)  # sample, weighted variance

xbar
s2
```

#### Trimmed Mean

```{r}
x <- c(0.32, 0.33, 0.36, 0.36, 0.39, 0.41, 0.49, 0.57, 0.6, 0.62, 0.63, 0.65, 0.65, 0.68, 0.72, 0.73, 0.76, 0.77, 0.78, 0.79)
mean(x, trim = .2)
```

## Dispersion

#### Range

The range is:

$$
\text{Range} = x_{\max} - x_{\min}
$$

```{r}
x <- c(30, 20, 10, 50, 100)

max_mass <- max(x, na.rm = TRUE)
min_mass <- min(x, na.rm = TRUE)

max_mass - min_mass
```

```{r}
range(data$variable, na.rm = TRUE)
range(x)
# Then max - min
```

#### Variance

For a population:

$$
\sigma^2 = \frac{\sum (x_i-\mu)^2}{N}
$$

where $\mu$ is the population mean and $N$ is the population size.

```{r}
# Population Variance
x <- c(24, 41, 58, 75, 92)

xbar <- mean(x)

variance <- sum((x - xbar)^2) / (length(x)) # use n
variance
```

For a sample:

$$
s^2 = \frac{\sum (x_i-\bar{x})^2}{n-1}
$$

where $\bar{x}$ is the sample mean and $n$ is the sample size.

```{r}
# Sample Variance
x <- c(24, 41, 58, 75, 92)

xbar <- mean(x)

variance <- sum((x - xbar)^2) / (length(x) - 1) # use n - 1
variance
```

```{r}
# Sample Variance Function
var(x)
```

#### MAD

```{r}
# Not tested.
```

#### Standard Deviation

Standard deviation is the square root of variance:

$$
s = \sqrt{s^2}
$$

For a population:

$$
\sigma = \sqrt{\frac{\sum (x_i-\mu)^2}{N}}
$$

```{r}
# Population Standard Deviation
x <- c(600, 470, 170, 430, 300)

xbar <- mean(x)
variance <- sum((x - xbar)^2) / length(x)
SD <- sqrt(variance)
SD
```

```{r}
# Sample Standard Deviation
x <- c(600, 470, 170, 430, 300)

xbar <- mean(x)
variance <- sum((x - xbar)^2) / (length(x) - 1)
SD <- sqrt(variance)
SD
```

```{r}
# Sample Standard Deviation Function
x <- c(600, 470, 170, 430, 300)

sd(x)
```

## Relative Positioning

#### z-scores

The z-score is:

$$
z = \frac{x-\mu}{\sigma}
$$

where $x$ is the observed value, $\mu$ is the mean, and $\sigma$ is the standard deviation.

```{r}
score <- 46.9
mu    <- 43.6
SD    <- 4.7

zscore <- (score - mu) / SD
zscore
```

#### Percentiles

```{r}
x <- c(6.1, 9.1, 9.5, 6, 8.6, 6.2, 9.1, 6.1, 8, 5.7, 6.5)

# One at a time
quantile(x, .28)

# Several at once
quantile(duration, c(.32, .57, .98))
```

#### Quartiles

```{r}
x <- c(10, 11, 13, 14, 14, 16, 17, 20, 25, 26, 26, 26, 27, 30, 33, 33, 33, 35, 38, 41, 41, 44, 44, 45, 45, 46)

x <- sort(x)
n <- length(x)

if (n %% 2 == 0) {
  lower <- x[1:(n/2)]
  upper <- x[(n/2 + 1):n]
} else {
  lower <- x[1:((n - 1) / 2)]
  upper <- x[((n + 3) / 2):n]
}


q1 <- median(lower)
q2 <- median(x)
q3 <- median(upper)

n # use as double-check

q1
q2
q3
```

```{r}
quantile(x, c(
  .25, # Q1
  .50, # Median
  .75  # Q3
  ), type = 2)
```

#### IQR

The interquartile range is:

$$
IQR = Q_3 - Q_1
$$

```{r}
# See calculations of quartiles above or manually set
q1 <- 205
q3 <- 248

IQR <- q3 - q1
IQR
```

```{r}
x <- c(51,8,3,35,7,18,23,7,9,52,39,2,23,32,43)
IQR(x, type = 2)
```

```{r}
IQR(data, type = 2)
```

## Two Variables

#### Covariance

The sample covariance is:

$$
s_{xy}
=
\frac{\sum (x_i-\bar{x})(y_i-\bar{y})}{n-1}
$$

A positive covariance indicates that the variables tend to move in the same direction, while a negative covariance indicates that they tend to move in opposite directions.

```{r}
x <- c(0.5, 1, 1.5, 2, 3, 3.5, 4, 4.5, 5.5, 6)
y <- c(63,	66,	69,	72,	78,	81,	84,	87,	90,	96)

# x
xbar <- mean(x)
xn   <- length(x)
xSD  <- sqrt( (sum((x - xbar)^2)) / (length(x)-1) ) # sample is n - 1

# y
ybar <- mean(y)
ySD  <- sqrt( (sum((y - ybar)^2)) / (length(y)-1) ) # sample is n - 1
yn   <- length(y)

# covariance
covariance <- (sum((x - xbar)*(y - ybar)) / (length(x)-1) )

covariance
```

```{r}
# using data
cov(x = data$x, y = data$y)

# using vectors
cov(x, y)
```

#### Correlation Coefficient

The correlation coefficient can be calculated from the covariance and standard deviations:

$$
r = \frac{s_{xy}}{s_xs_y}
$$

where $s_{xy}$ is the sample covariance and $s_x$ and $s_y$ are the sample standard deviations.

```{r}
# REQUIRED: Use the covariance section above AND run it !
# Then, calculate r
r <- covariance / (xSD * ySD)
r
```

```{r}
# using data
cor(x = data$x, y = data$y)

# using vectors
cor(x, y)
```
