---
title: "exam3"
format: html
execute:
  eval: false
  warning: false
  message: false
---

## Additional Helpful Functions

-   `janitor::round_half_up(calculation, round)`

-   `round(calculation, digits = round)`

## Uniform Distribution

### Identification

Problem gives:

-   the variable is uniformly distributed

-   values lie between $a$ and $b$

-   every value in the interval is equally likely

### Key formulas

Mean:

$$
\mu = \frac{a+b}{2}
$$

Standard deviation

$$
\sigma = \sqrt{\frac{(b-a)^2}{12}}
$$

Probability to the right of a cutoff:

$$
P(X > x) = \frac{b-x}{b-a}
$$

Probability to the left of a cutoff:

$$
P(X < x) = \frac{x-a}{b-a}
$$

### Example script

```{r}
# mean and standard deviation
mu <- (a + b) / 2
sigma <- sqrt((b - a)^2 / 12)

# probability greater than x
(b - x) / (b - a)

# probability less than x
(x - a) / (b - a)
```

## Normal distribution with a single raw score

### Identification

Problem gives:

-   a mean and standard deviation

-   says that the variable is normally distributed

-   asks about a randomly selected observation

-   uses $X$ rather than $\bar{x}$

### Key formulas

For a value between two numbers:

$$
P(a < X < b)
$$

Standardizing:

$$
P(a < X < b)
$$

For a value above a cutoff:

$$
P(X > x)
$$

For a value below a cutoff:

$$
P(X < x)
$$

For finding a cutoff from a percentile:

$$
P(X \le x) = p
$$

For solving for standard deviation:

$$
z = \frac{x - \mu}{\sigma}\qquad \Rightarrow \qquad\sigma = \frac{x - \mu}{z}
$$

### Example script 

```{r}
# between two values
pnorm(upper, mean = mu, sd = sigma) - 
  pnorm(lower, mean = mu, sd = sigma)

# greater than
1 - pnorm(x, mean = mu, sd = sigma)

# less than
pnorm(x, mean = mu, sd = sigma)

# cutoff from a percentile
qnorm(p, mean = mu, sd = sigma)

# solve for sigma
z <- qnorm(p_left)
sigma <- (x - mu) / z
```

## Graphs of Normal Distributions

### Identification

Conceptual problems comparing:

-   centers of the curves

-   width/spread of the curves

-   whether means are the same or different

-   whether standard deviations are the same or different

### Key ideas

$$
\text{Different centers} \Rightarrow \text{different means}\\
\text{Wider/flatter curve} \Rightarrow \text{larger standard deviation}\\
\text{Narrower/taller curve} \Rightarrow \text{smaller standard deviation}
$$

## Empirical Rule

### Identification

Problem shows:

-   a normal curve

-   shaded regions

-   labels involving 𝜇, 𝜇 + 𝜎, 𝜇 + 2𝜎, 𝜇 + 3𝜎

### Key ideas

$$
68\% \text{ is within } \mu \pm \sigma\\
95\% \text{ is within } \mu \pm 2\sigma\\
99.7\% \text{ is within } \mu \pm 3\sigma
$$

### Supporting Script

```{r}
curve(dnorm(x), from = -4, to = 4)
abline(v = c(0, 1, 2, 3), lty = 2)
```

## Sampling distribution of the mean $\bar{X}$

### Identification

Problem gives:

-   “sample mean”

-   “a random sample of $n$”

-   “what is the probability that the sample mean...”

-   “sampling distribution”

### Key formulas

Mean of the sampling distribution:

$$
\mu_{\bar{X}} = \mu
$$

Standard deviation of the sampling distribution, also called the standard error:

$$
\sigma_{\bar{X}} = \frac{\sigma}{\sqrt{n}}
$$

If variance is given first:

$$
\sigma = \sqrt{\text{variance}}\\
\sigma_{\bar{X}} = \frac{\sigma}{\sqrt{n}}
$$

Probability forms:

$$
P(\bar{X} < x)\\
P(\bar{X} > x)\\
P\bigl(|\bar{X} - \mu| < a\bigr)\\
P\bigl(|\bar{X} - \mu| > a\bigr)
$$

Expanded interval forms:

$$
P(\mu - a < \bar{X} < \mu + a)\\
P(\bar{X} < \mu - a) + P(\bar{X} > \mu + a)
$$

### Example script

```{r}
# if variance is given
sigma <- sqrt(variance)

# standard error
se <- sigma / sqrt(n)

# less than
pnorm(xbar, mean = mu, sd = se)

# greater than
1 - pnorm(xbar, mean = mu, sd = se)

# within a distance of mu
pnorm(mu + amount, mean = mu, sd = se) -
  pnorm(mu - amount, mean = mu, sd = se)

# greater than a distance from mu
1 - (pnorm(mu + amount, mean = mu, sd = se) -
       pnorm(mu - amount, mean = mu, sd = se))
```
