---
title: "Exam 2 Review"
format: html
execute:
  eval: false
  warning: false
  message: false
---

## Additional helpful functions

-   `janitor::round_half_up(calculation, round)`

-   `MASS::fractions(calculation)`

| Problem                 | Phrase                          |
|-------------------------|---------------------------------|
| Sample space            | list all outcomes               |
| Basic probability       | favorable / total               |
| Mutually exclusive      | Check overlap                   |
| Conditional probability | "given that..."                 |
| Multiplication rule     | independent events              |
| Dependent events        | without replacement             |
| Combinations            | selecting groups                |
| Complement rule         | "at least... "                  |
| Expected value          | gains/losses with probabilities |
| Binomial                | fixed trials, two outcomes      |
| Poisson                 | count of events in an interval  |

## Probability

$$
P(A) = \frac{\text{number of favorable outcomes}}{\text{total number of outcomes}}
$$

```{r}

```

### Unions (A or B)

$$
P(A \cup B) = P(A) + P(B) - P(A \cap B)
$$

```{r}

```

### Conditional Probability (A given B)

$$
P(A \mid B) = \frac{P(A \cap B)}{P(B)}
$$

```{r}

```

### Intersections (A and B)

#### Independence

$$
P(A \cap B) = P (A\;and\;B) = P(A) \times P(B)
$$

```{r}

```

#### Dependent (Multiplication Rule)

$$
P(A \cap B) = P (A\;and\;B)= P(A) \times P(B \vert A)
$$

```{r}

```

### Complement

$$
P(A^c)= P (not\;A) = 1 - P(A)
$$

```{r}

```

## Discrete Random Variable

#### Expected Value

$$
E(X) = \sum_{i} x_i \, P(X = x_i)
$$

```{r}

```

#### Variance and Standard Deviation

$$
\mathrm{Var}(X) = \sum_{i} (x_i - \mu)^2 \, P(X = x_i)\\\sigma = \sqrt{\mathrm{Var}(X)}
$$

```{r}

```

## Discrete Uniform Distribution

$$
 P(X = x) = \frac{1}{n}
$$

```{r}

```

#### Expected Value

$$
E(X) = \frac{\text{first value} + \text{last value}}{2}
$$

```{r}

```

## Binomial Distribution

$$
 P(X = x) = \binom{n}{x} p^x (1-p)^{n-x}
$$

+----------------+-------------------+-------------------+
| Wording        | Mathematical Form | R Function        |
+================+===================+===================+
| exactly x      | $P(X=x)$          | `dbinom()`        |
+----------------+-------------------+-------------------+
| at most x\     | $P(X\le x)$       | `pbinom()`        |
| no more than x |                   |                   |
+----------------+-------------------+-------------------+
| less than x    | $P(X \lt x)$      | `pbinom(x-1)`     |
+----------------+-------------------+-------------------+
| more than x    | $P(X \gt x)$      | `1 - pbinom(x)`   |
+----------------+-------------------+-------------------+
| at least x     | $P(X \ge x)$      | `1 - pbinom(x-1)` |
+----------------+-------------------+-------------------+

| Argument | Equation | Meaning                             |
|----------|----------|-------------------------------------|
| `x`      | $x$      | number of successes                 |
| `size`   | $n$      | number of trials                    |
| `prob`   | $p$      | probabiity of success on each trial |

```{r}

```

#### Expected Value

$$
E(X) = np
$$

```{r}

```

#### Variance and Standard Deviation

$$
Var(X) = np(1-p)\\SD(X) = \sqrt{np(1-p)}
$$

```{r}

```

## Poisson Distribution

$$
P(X = x) = \frac{e^{-\lambda}\lambda^x}{x!}, \quad x = 0,1,2,\ldots
$$

+----------------+-------------------+------------------+
| Wording        | Mathematical Form | R Function       |
+================+===================+==================+
| exactly x      | $P(X=x)$          | `dpois()`        |
+----------------+-------------------+------------------+
| at most x\     | $P(X\le x)$       | `ppois()`        |
| no more than x |                   |                  |
+----------------+-------------------+------------------+
| less than x    | $P(X \lt x)$      | `ppois(x-1)`     |
+----------------+-------------------+------------------+
| more than x    | $P(X \gt x)$      | `1 - ppois(x)`   |
+----------------+-------------------+------------------+
| at least x     | $P(X \ge x)$      | `1 - ppois(x-1)` |
+----------------+-------------------+------------------+

| Argument | Equation  | Meaning |
|----------|-----------|---------|
| `x`      | $x$       | vector  |
| `lambda` | $\lambda$ | lambda  |

```{r}

```

## Hypergeometric Distribution

$$
P(X = x) = \frac{\binom{k}{x}\binom{N-k}{\,n-x\,}}{\binom{N}{n}}
$$

+----------------+-------------------+-------------------+
| Wording        | Mathematical Form | R Function        |
+================+===================+===================+
| exactly x      | $P(X=x)$          | `dhyper()`        |
+----------------+-------------------+-------------------+
| at most x\     | $P(X\le x)$       | `phyper()`        |
| no more than x |                   |                   |
+----------------+-------------------+-------------------+
| less than x    | $P(X \lt x)$      | `phyper(x-1)`     |
+----------------+-------------------+-------------------+
| more than x    | $P(X \gt x)$      | `1 - phyper(x)`   |
+----------------+-------------------+-------------------+
| at least x     | $P(X \ge x)$      | `1 - phyper(x-1)` |
+----------------+-------------------+-------------------+

| Argument | Equation | Meaning                 |
|----------|----------|-------------------------|
| `x`      | $x$      | successes in sample     |
| `m`      | $k$      | successes in population |
| `n`      | $N-k$    | failures in population  |
| `k`      | $n$      | sample size             |

```{r}

```

#### Expected Value and Variance

$$
E(X) = n \cdot \frac{k}{N}
$$

```{r}

```

#### Variance

$$
\operatorname{Var}(X)=n \cdot \frac{k}{N}\left(1 - \frac{k}{N}\right)\frac{N - n}{N - 1}
$$

```{r}

```
