---
title: "Exam 1 Formula Guide"
format: 
  html: 
    toc: true
    toc-location: left
    toc-title: "Questions"
    toc-depth: 2
    css: styles.css
engine: knitr
execute:
  eval: false
  warning: false
  message: false
---

## Suggestions

-   Classify the problem to determine which calculation or formula you need.

-   Careful writing down the data! Watch out for [typos]{.underline}.\
    Pasting tables into Excel might paste pictures (whomp whomp).

-   Round to the specified number of digits [at the last calculation]{.underline}.

-   Be [strategic]{.underline} with your available time.

-   **YOU GOT THIS!**

### Helpful Excel Formulas

-   =TEXTJOIN(", ", TRUE, A1:A14)

-   =SUM()

-   Simple math ( + - / etc.)

-   Lock a cell reference with \$ (e.g., \$A\$1)

## Frequency Tables

```{r}
library(descr)
```

```{r}
descr::freq(as.ordered(data$variable), plot = FALSE)
```

## Central Tendency

#### Mode

```{r}
library(lsr)
```

```{r}
lsr::modeOf(data$variable, na.rm = TRUE)
```

```{r}
# Simple histogram
x <- c(13,13,-5,-5,13,11,-5)
hist(x, breaks = 10) # breaks = bins = classes
```

#### Median

```{r}
x <- c(13,13,-5,-5,13,11,-5)
median(x)
```

```{r}
median(data$variable, na.rm = TRUE)
```

#### Mean

```{r}
(56 + 31 + 56 + 8 + 32) / 5
```

```{r}
x <- c(13,13,-5,-5,13,11,-5)
sum(x) / length(x)
```

```{r}
x <- c(56, 31, 56, 8, 32)
mean(x, na.rm = TRUE)
```

#### Weighted Mean

```{r}
x <- c(24, 41, 58, 75, 92) # observations
w <- c(2, 8, 15, 8, 11)  # weights

sum(x * w) / sum(w)
```

```{r}
x <- c(24, 41, 58, 75, 92) # observations
w <- c(2, 8, 15, 8, 11)  # weights

weighted.mean(x, w, na.rm = TRUE)
```

```{r}
# variance, calculate after calculating the mean
x <- c(24, 41, 58, 75, 92) # observations
w <- c(2, 8, 15, 8, 11)  # weights

n <- sum(w)

xbar <- sum(x * w) / n
s2   <- sum(w * (x - xbar)^2) / (n - 1)

xbar
s2
```

#### Trimmed Mean

```{r}
x <- c(0.32, 0.33, 0.36, 0.36, 0.39, 0.41, 0.49, 0.57, 0.6, 0.62, 0.63, 0.65, 0.65, 0.68, 0.72, 0.73, 0.76, 0.77, 0.78, 0.79)
mean(x, trim = .2)
```

## Dispersion

#### Range

```{r}
x <- c(30, 20, 10, 50, 100)

max_mass <- max(x, na.rm = TRUE)
min_mass <- min(x, na.rm = TRUE)

max_mass
min_mass
```

```{r}
range(data$variable, na.rm = TRUE)
```

#### Variance

```{r}
x <- c(24, 41, 58, 75, 92)

xbar <- mean(x)
variance <- sum((x - xbar)^2) / (length(x) - 1)

variance
```

#### MAD

```{r}

```

#### Standard Deviation

```{r}
x <- c(600, 470, 170, 430, 300)

xbar <- mean(x)
variance <- sum((x - xbar)^2) / length(x)
SD <- sqrt(variance)

SD
```

## Relative Positioning

#### z-scores

```{r}
score <- 46.9
mu    <- 43.6
SD    <- 4.7

zscore <- (score - mu) / SD
zscore
```

#### Percentiles

```{r}
summary(data)
```

```{r}
x <- c(6.1, 9.1, 9.5, 6, 8.6, 6.2, 9.1, 6.1, 8, 5.7, 6.5, 6.4, 5.8, 9.3, 6.2
)

quantile(x, .28)
```

```{r}
quantile(duration, c(.32, .57, .98))
```

#### Quartiles

```{r}
x <- c(10, 11, 13, 14, 14, 16, 17, 20, 25, 26, 26, 26, 27, 30, 33, 33, 33, 35, 38, 41, 41, 44, 44, 45, 45, 46)

x <- sort(x)
n <- length(x)

if (n %% 2 == 0) {
  lower <- x[1:(n/2)]
  upper <- x[(n/2 + 1):n]
} else {
  lower <- x[1:((n - 1) / 2)]
  upper <- x[((n + 3) / 2):n]
}


q1 <- median(lower)
q2 <- median(x)
q3 <- median(upper)

n # use as double-check

q1
q2
q3
```

```{r}
quantile(x, c(
  .25, # Q1
  .50, # Median
  .75  # Q3
  ), type = 2)
```

#### IQR

```{r}
# See calculations of quartiles above or manually set
q1 <- 205
q3 <- 248

IQR <- q3 - q1
IQR
```

```{r}
x <- c(51,8,3,35,7,18,23,7,9,52,39,2,23,32,43)
IQR(x, type = 2)
```

```{r}
IQR(data, type = 2)
```

## Two Variables

#### Covariance

```{r}
x <- c(0.5, 1, 1.5, 2, 3, 3.5, 4, 4.5, 5.5, 6)
y <- c(63,	66,	69,	72,	78,	81,	84,	87,	90,	96)

# x
xbar <- mean(x)
xn   <- length(x)
xSD  <- sqrt( (sum((x - xbar)^2)) / (length(x)-1) ) # sample is n - 1

# y
ybar <- mean(y)
ySD  <- sqrt( (sum((y - ybar)^2)) / (length(y)-1) ) # sample is n - 1
yn   <- length(y)

# covariance
covariance <- (sum((x - xbar)*(y - ybar)) / (length(x)-1) )

covariance
```

```{r}
# cov(x = data$x, y = data$y)
```

#### Correlation Coefficient

```{r}
# REQUIRED: Update the covariance section above AND run it ! 
r <- covariance / (xSD * ySD)

r
```

```{r}
cor(x = data$x, y = data$y)
```
